Collective excitations in competing phases in two and three dimensions

### 2024-03-06 paper/main.tex
mm

@@

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/phase_diagram.pdf}
    \caption{The phase diagram obtained for the extended Hubbard model on 
    (a) a square lattice and (b) a simple cubic lattice using a mean-field approximation at temperature $T=0$.
    The CDW-AFM boundary lies at $U = zV$, where $z$ is the coordination number. 
    For $U<0$ and $V=0$, there is a coexistence of CDW and $s$-wave SC.
    The boundary for the $d_{x^2 - y^2}$-wave SC (dashed line) for the square lattice has been taken from Ref. \cite{Micnas88b}, 
    as it is not accessible using our formalism.
    \textcolor{red}{The red shading shows the region in which the dynamical matrix $\mM$, see equation \eqref{eqn:heisenberg}, has negative eigenvalues, 
    indicating that there are phases not captured by our methods.
    In accordance with the literature, we expect to find a phase-separated state \cite{Linner23}.
    Lastly, the red shading shows the region that we cannot access because the discretization $N_\gamma$, 
    which is required to accurately describe the system, is too large.}}
    \label{fig:phase_diagram}
\end{figure}

@@

Naturally, we have to discretize $\gamma$, 
however, this allows us to obtain good results with drastically smaller matrices compared to a momentum-based operator set and obtain results for a three-dimensional system in the first place.
Practically, we choose $N_\gamma = 6000$ equally spaced discretization points.
A detailed explanation of the numerics is provided in the appendix \ref{sec:numerical_ieom}.
\textcolor{red}{Computationally, our method is limited by the number of terms included in our operator set as well in the magnitude of $N_\gamma$.}
If the gap $\Delta_\text{tot}$ is of the same order of magnitude as our discretization $\Delta \gamma$ our numerics become inaccurate.
This manifests specifically if a spectral function has strong features at minuscule energies, e.g., the phase mode in the superconducting phase.
In this case, the dynamical matrix $\mM$ picks up negative eigenvalues that vanish if $N_\gamma$ is increased so that $\Delta \gamma$ is sufficiently smaller than $\Delta_\text{tot}$.
This happens if the parameters are chosen from the purple shaded area in \autoref{fig:phase_diagram}.
\textcolor{red}{This issue only arises in the SC phase, nevertheless, we expect similar inaccuracies in the AFM region for small $U$.
However, we did not perform any computations in that region, hence this does not pose an issue to our analysis.}

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/afm_cdw_peaks_overview.pdf}
    \caption{\textcolor{red}{The upper panels (a) show plots of the positions $\omega_0$ of the peak in $\spectral{CDW}$ (orange lines) and $\spectral{l.AFM}$ (blue lines) close to the corresponding phase transition $U = 4.8t$.
    Again, we chose $V=1.2t$ for the square lattice (left column) and $V=0.8t$ for the simple cubic lattice (right column) respectively.
    The middle panels (b) depict the peak positions relative to the lower edge of the two-particle continuum $\omega_-$, while the bottom panels (c) show their respective weights.}
    The shadings indicate the phase the system is in. Blue represents AFM while orange represents CDW.
    Note the difference in the scales for the two lattices.}
    \label{fig:afm_cdw_peaks_overview}
\end{figure*}

@@

Note, that all of the aforementioned operators are Hermitian. 
This means that all spectral functions need to be completely antisymmetrical \cite{rickayzen80}.%, see appendix \ref{sec:antisymmetry_spectral}.
For this reason, we will only show the $\omega \geq 0$ in our plots. 
Additionally, we will add a small imaginary part of $10^{-5}t$ to $\omega$ in order to visualize the occurring $\delta$ peaks.

@@

We show that $\mathcal{M}$ is positive semidefinite if the system is in thermal equilibrium.
For any $\vec{x} \in \mathbb{C}^N$, consider
\begin{align}
    \vec{x}^\dagger \mM \vec{x} &= \sum_{ij} x_i^* x_j \mM_{ij} \nonumber \\
        &= \langle \left[ \sum_i x_i^* A_i^\dagger, \left[ H, \sum_j x_j A_j \right] \right]  \rangle \nonumber \\
        &= \langle [B^\dagger, [H, B]] \rangle\,,\quad B \coloneqq  \sum_j x_j A_j\,.
\end{align}
For $\mM$ to be positive semidefinite the expression above must not be negative.
Therefore, we need to show, that any diagonal element of $\mM$ is nonnegative.
To do this, we expand the commutators and the thermal expectation value $\langle O \rangle = \frac{1}{Z} \Trace [\exp(-\beta H_\text{MF}) O]$, 
where $\beta$ is the inverse temperature and $Z$ the partition function.
This procedure yields

\begin{widetext}
    \begin{align}
        \langle [B^\dagger, [H, B]] \rangle &= \frac{1}{Z} \Trace [\exp(-\beta H) [B^\dagger, [H, B]]] \nonumber \\
            &= \frac{1}{Z} \sum_j e^{-\beta \epsilon_j} \bra{\psi_j} [B^\dagger, [H, B]] \ket{\psi_j} \nonumber \\
            &= \frac{1}{Z} \sum_j e^{-\beta \epsilon_j} \left[ \bra{\psi_j} B^\dagger (H - \epsilon_j) B \ket{\psi_j} + \bra{\psi_j} B (H - \epsilon_j) B^\dagger \ket{\psi_j} \right] \nonumber \\
            &= \frac{1}{Z} \sum_{ij} e^{-\beta \epsilon_j} \left[ \bra{\psi_j} B^\dagger \ket{\psi_i} (\epsilon_i - \epsilon_j) \bra{\psi_i} B \ket{\psi_j} + \bra{\psi_j} B \ket{\psi_i} (\epsilon_i - \epsilon_j) \bra{\psi_i} B^\dagger \ket{\psi_j} \right] \nonumber \\
            &= \frac{1}{Z} \sum_{ij} e^{-\beta \epsilon_j} (\epsilon_i - \epsilon_j) \underbrace{ \left[ \bra{\psi_j} B^\dagger \ket{\psi_i} \bra{\psi_i} B \ket{\psi_j} + \bra{\psi_j} B \ket{\psi_i} \bra{\psi_i} B^\dagger \ket{\psi_j} \right]}_{\coloneqq C_{ij} = C_{ji} \geq 0} \nonumber \\
            &= \frac{1}{2Z} \left[ \sum_{ij} C_{ij} e^{-\beta \epsilon_j} (\epsilon_i - \epsilon_j) - \sum_{ij} C_{ij} e^{-\beta \epsilon_i} (\epsilon_i - \epsilon_j) \right] \nonumber \\
            &= \frac{1}{2Z} \sum_{ij} C_{ij} \left( e^{-\beta \epsilon_j} - e^{-\beta \epsilon_i} \right) \left( \epsilon_i - \epsilon_j \right) \geq 0 \,.
    \end{align}
\end{widetext}

This concludes our proof. Note, that the only assumption made, is that the expectation value is to be taken with respect to the thermal equilibrium.

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/higgs_peak.pdf}
    \caption{Log-log plot of $\spectral{Higgs}$ (a, c) and $\spectral{CDW}$ (b) near $\omega = \omega_0 = 2\Delta_\text{SC}$ at $U=-2.5t$ and $V=0$.
    The different panels depict the spectral functions for various ratios of $\Delta_\text{SC}$ to $\Delta_\text{CDW}$.
    We additionally plot the difference between $\spectral{i}$ and $\spectral{Phase}$ and fit it, as we interpret the latter as a background within the two-particle continuum.
    The left (right) panel shows the results for the square (simple cubic) lattice.
    The spectral function behaves as $1 / \sqrt{\omega - 2 \Delta_\text{tot}}$.
    Note, that $\spectral{CDW}$ and $\spectral{Higgs}$ are identical for the choice in panel (c).}
    \label{fig:higgs_peak}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=0.98\textwidth]{plots/peak_fits.pdf}
    \caption{Log-log plot of the real part of various Green's functions, the labels indicate its type and the text in the panels indicate the phase.
        The results for both lattice types are shown but are qualitatively identical.
        The power-law exponent is $a=-1$ in all cases, indicating that these peaks are $\delta$ distributions.}
        \label{fig:peak_fits}
\end{figure*}

In this appendix section, we show fits to the real and imaginary parts of various Green's functions in the different phases.
All fits are linear of the type $y = ax + b$ in double logarithmic plots, i.e., $a$ describes the power-law behavior of the functions.
Firstly, the real part of $\greens{Phase}$ in the SC phase is shown in panel (a) of \autoref{fig:zero_peaks}.
It behaves as $1/\omega^2$ indicating that the peak we find in the spectral function at $\omega=0$ corresponds to the derivative of a $\delta$ distribution.
The same results are obtained for the peak in $\spectral{Higgs}$ and $\spectral{CDW}$ in the coexistence phase.

Continuing, we investigate the divergence at $\omega = 2\Delta$ found in $\spectral{Higgs}$ and $\spectral{CDW}$ at $V=0$.
It behaves as $1/\sqrt{\omega}$ as shown in \autoref{fig:higgs_peak}.

Lastly, we show all the remaining peaks in \autoref{fig:peak_fits}.
The real part of each one of them behaves as $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The peaks of $\spectral{AFM}$ and $\spectral{CDW}$ close the the phase transition are identical when swapped, i.e., $\spectral{AFM}$ in the CDW phase is the same as $\spectral{CDW}$ in the AFM phase.

\section{Numerical treatment of the iEoM matrices}
\label{sec:numerical_ieom}

@@

@@

>>>>

\def\hmath$#1${\texorpdfstring{{\rmfamily\textit{#1}}}{#1}}

@@

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/phase_diagram.pdf}
    \caption{The phase diagram obtained for the extended Hubbard model on 
    (a) a square lattice and (b) a simple cubic lattice using a mean-field approximation at temperature $T=0$.
    The CDW-AFM boundary lies at $U = zV$, where $z$ is the coordination number. 
    For $U<0$ and $V=0$, there is a coexistence of CDW and $s$-wave SC.
    The boundary for the $d_{x^2 - y^2}$-wave SC (dashed line) for the square lattice has been taken from Ref. \cite{Micnas88b}, 
    as it is not accessible using our formalism.
    \textcolor{red}{The red shading shows the region in which the dynamical matrix $\mM$, see equation \eqref{eqn:heisenberg}, has negative eigenvalues, 
    indicating that there are phases not captured by our methods.
    In accordance with the literature, we expect to find a phase-separated state \cite{Linner23}.
    Lastly, the striped area shows the region that we cannot access because the discretization $N_\gamma$, 
    which is required to accurately describe the system, is too large. However, we expect to find a superconducting state here.}}
    \label{fig:phase_diagram}
\end{figure}

@@

Naturally, we have to discretize $\gamma$, 
however, this allows us to obtain good results with drastically smaller matrices compared to a momentum-based operator set and obtain results for a three-dimensional system in the first place.
Practically, we choose $N_\gamma = 6000$ equally spaced discretization points.
A detailed explanation of the numerics is provided in the appendix \ref{sec:numerical_ieom}.
\textcolor{red}{Computationally, our method is limited by the number of terms included in our operator set as well in the magnitude of $N_\gamma$.}
If the gap $\Delta_\text{tot}$ is of the same order of magnitude as our discretization $\Delta \gamma$ our numerics become inaccurate.
This manifests specifically if a spectral function has strong features at minuscule energies, e.g., the phase mode in the superconducting phase.
In this case, the dynamical matrix $\mM$ picks up negative eigenvalues that vanish if $N_\gamma$ is increased so that $\Delta \gamma$ is sufficiently smaller than $\Delta_\text{tot}$.
This happens if the parameters are chosen from the striped area in \autoref{fig:phase_diagram}.
\textcolor{red}{This issue only arises in the SC phase, nevertheless, we expect similar inaccuracies in the AFM region for small $U$.
However, we did not perform any computations in that region, hence this does not pose an issue to our analysis.}

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/afm_cdw_peaks_overview.pdf}
    \caption{\textcolor{red}{The upper panels (a) show plots of the positions $\omega_0$ of the peak in $\spectral{CDW}$ (orange lines) and $\spectral{l.AFM}$ (blue lines) close to the corresponding phase transition $U = 4.8t$.
    Again, we chose $V=1.2t$ for the square lattice (left column) and $V=0.8t$ for the simple cubic lattice (right column) respectively.
    The middle panels (b) depict the peak positions relative to the lower edge of the two-particle continuum $\omega_-$, while the bottom panels (c) show their respective weights.}
    The shadings indicate the phase the system is in. Green represents AFM while orange represents CDW.
    Note the difference in the scales for the two lattices.}
    \label{fig:afm_cdw_peaks_overview}
\end{figure*}

@@

Note, that all of the aforementioned operators are Hermitian. 
This means that all spectral functions need to be completely antisymmetrical \cite{rickayzen80}.%, see appendix \ref{sec:antisymmetry_spectral}.
For this reason, we will only show the $\omega \geq 0$ in our plots. 
Additionally, we will add a small imaginary part of $10^{-5}t$ to $\omega$ in order to visualize the occurring $\delta$ peaks.
\textcolor{red}{Furthermore, we provide a brief analysis of the effect of choosing different numbers of discretization points $N_\gamma$ in appendix \ref{sec:finite_size}.}

@@

We show that $\mathcal{M}$ is positive semidefinite if the system is in thermal equilibrium.
For any $\vec{x} \in \mathbb{C}^N$, consider
\begin{align}
    \vec{x}^\dagger \mM \vec{x} &= \sum_{ij} x_i^* x_j \mM_{ij} \nonumber \\
        &= \langle \left[ \sum_i x_i^* A_i^\dagger, \left[ H, \sum_j x_j A_j \right] \right]  \rangle \nonumber \\
        &= \langle [B^\dagger, [H, B]] \rangle\,,\quad B \coloneqq  \sum_j x_j A_j\,.
\end{align}
For $\mM$ to be positive semidefinite the expression above must not be negative,
\textcolor{red}{which has already been proven in Refs. \cite{Dyson1978,mermin66}.}

@@

In this appendix section, we discuss fits to the real and imaginary parts of various Green's functions in the different phases.
All fits are linear of the type $y = ax + b$ in double logarithmic plots, i.e., $a$ describes the power-law behavior of the functions.
An example plot of the real part of $\greens{Phase}$ in the SC phase is shown in panel (a) of \autoref{fig:zero_peaks}.
It behaves as $1/\omega^2$ indicating that the peak we find in the spectral function at $\omega=0$ corresponds to the derivative of a $\delta$ distribution.
The same results are obtained for the peak in $\spectral{Higgs}$ and $\spectral{CDW}$ in the coexistence phase as well as in $\spectral{t.AFM}$ within the AFM phase.

Continuing, we investigate the divergence at $\omega = 2\Delta$ found in $\spectral{Higgs}$ and $\spectral{CDW}$ at $V=0$. It behaves as $1/\sqrt{\omega}$.

Lastly, all remaining peaks below the two-particle continuum behave identical.
The real part of each one of them behaves as $1/\omega$ in close vicinity to the peak position.
This indicates that the peaks are $\delta$ distributions.
The peaks of $\spectral{AFM}$ and $\spectral{CDW}$ close the the phase transition are identical when swapped, 
i.e., $\spectral{AFM}$ in the CDW phase is the same as $\spectral{CDW}$ in the AFM phase.

\section{Numerical treatment of the iEoM matrices in \hmath $\gamma$-space}
\label{sec:numerical_ieom}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{plots/finite_size.pdf}
    \caption{Effect of changing the number of discretization points $N_\gamma$ on the absolute position of the peak in $\spectral{Phase}$ for $U=-2.5t$ and $V=-0.1t$.
    On the simple cubic lattice, the peak already lies properly at $\omega=0$.
    On the square lattice, its position approaches $0$ with $\sqrt{1/N_\gamma}$.}
    \label{fig:finite_size}
\end{figure}

@@

\section{Dependence of the results on the discretization}
\label{sec:finite_size}

Most of our data barely depends on the number of discretization points $N_\gamma$ at all.
However, specifically features that lie at very small energies, e.g., the peak in $\spectral{Phase}$ within the SC phase, can be difficult to properly resolve.
We show this effect in \autoref{fig:finite_size}.
Neglecting minor fluctuations, this peak is properly located at $\omega=0$ for the simple cubic lattice.
On the square lattice, however, it follows a $\sqrt{1/N_\gamma}$ behavior, i.e., also approaching $0$.
This behavior can be observed for the other peaks at $\omega=0$ as well.
How far away they are from $0$, or whether they are at all, depends on the spectral function as well as on the system parameters.
